Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-02-28 application.tex
minor changes
redoing application section

In our disk models we have adopted a simple thermal relaxation 
function in the energy equation. To connect our model with a realistic
disk, we consider a more physically-motivated form of the energy
equation with radiative cooling 
\begin{align}\label{real_energy}
\rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}
where the temperature $T$ is defined through $P=\mathcal{R}\rho
T/\mu$, $\mathcal{R}$ is the ideal gas constant and $\mu$ is
the mean molecular weight, % In Eq. \ref{real_energy} $S$ is the
% specific entropy,
% \begin{align}
%   S \equiv c_v\ln{\left(\frac{p}{\rho^{\gamma}}\right)} 
% \end{align}
% where $c_v \equiv \left(\gamma-1\right)^{-1}\mathcal{R}/\mu$ is the
% specific heat capcity at constant volume
$D/Dt$ is the convective
derivative, $\bm{F}$ is the radiative flux, and $Q_+$ is a heating
term such that an equilibrium solution exists when $D/Dt=0$.

Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
the thermal relaxation time $t_c$ as used in our linear
calculations. To do so, we first consider separately perturbations
with radial lengthscales $l\equiv 2\pi/k_x\ll l_\mathrm{rad}$ and 
$l\gg l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity. 
%need to comment on kappa_d(rho,T)

\subsection{Newtonian cooling}\label{newton_cool}
For $l\ll l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right), 
\end{align}
where $c_v$ is the heat capacity at constant volume. When linearized, the last term on the RHS becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{full_energy1}), we identify 
\begin{align}\label{tc_newton_cool} 
  t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
\end{align}
as the thermal relaxation timescale for perturbation lengthscales
$l\ll l_\mathrm{rad}$.

\subsection{Radiative diffusion}
For perturbation lengthscales $l\gg l_\mathrm{rad}$, we assume 
temperature fluctuations are smoothed out by radiative diffusion. To
model this in linear theory, we choose the heating term such that
\begin{align}
  Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
    \left(T-T_{t=0}\right)\right], 
\end{align} 
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly. Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
    c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right). 
\end{align}
This, of course, fundamentally changes the linear problem by
introducing higher order derivatives. We defer a proper analysis of
this problem to a follow-up study. Here, we are interested in
order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed
by assuming vertical derivatives can be neglected in
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
% we are interested in radially-thin perturbations  
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}. 
\end{align}
From Eq. \ref{diff_cool_approx} we identify the thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}
where $\hat{\eta} = \eta/H_\mathrm{iso}^2\Omega_k$ is the 
dimensionless diffusion coefficient.

\subsection{Toy model for realistic thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff} =
  \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
$\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
and Eq. \ref{eta_def}. Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$.

In order to utilize our linear results, which assumes $t_c$ is a
constant, we need to make further 
approximations. Specifically, we consider vertically-isothermal disks
and replace explicit $\rho$ dependencies in the above expressions by $\Sigma/
H_\mathrm{iso}$, where $\Sigma$ is the surface density at the fiducial
radius of interest. For example, the dimensionless diffusion
coefficient becomes 
\begin{align}\label{diff_coeff_dimensionless}
  \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
    c_v\Omega_k}. 
\end{align}
Furthermore, we will choose a temperature-dependent opacity model for
simplicity, so that $\kappa_d=\kappa_d(T)$. The dimensionless thermal
relaxation time $\beta$  
\begin{align}\label{real_beta}
  \beta \equiv t_c\Omega_k =
  \frac{1}{\hat{\eta}}\left(\frac{l_\mathrm{rad}^2}{3H_\mathrm{iso}^2}
    + \frac{1}{\khat^2}\right)  \simeq
  \frac{1}{\hat{\eta}}\left(\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{1}{\khat^2}\right)
\end{align}
is then a constant for vertically-isothermal disks, 
and we can directly apply previous results.

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).    
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\zeta^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is necessary for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
%becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} for the Minimum Mass Solar Nebula (MMSN) by setting
the scaling $\hat{T}=\hat{\Sigma}=\hat{\kappa}_d=1$, with $\mu =
2.33$ and $\gamma=1.4$. The figure suggests that the fundamental VSI
operates in the outer parts of the MMSN at radii on 
the order of tens of AU. However, the thermal relaxation time $\beta$
is not much shorter than $\beta_\mathrm{crit}$, so the VSI may be 
weak.

We remark that it is possible to satisfy $\beta<\beta_\mathrm{crit}$
at small radii by choosing a sufficiently large $\khat$. However, our
linear calculations suggest growth rates eventually decay as 
$\khat\to\infty$. Thus, very small-scale VSI in the inner disk is
probably unimportant.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/bcrit_mmsn}  
  \caption{Dimensionless thermal relaxation timescales $\beta$ in the MMSN
    using Eq. \ref{beta_mmsn} for three values of the 
    perturbation radial wavenumber: $\khat=1$ (dotted), $\khat=10$
    (dashed) and $\khat=100$ (dashed-dot). The solid line is the
    critical thermal relaxation timescale $\beta_\mathrm{crit}$, and 
    $\beta<\beta_\mathrm{crit}$
    is necessary for the fundamental isothermal VSI to operate  
    for $\khat\gg1$. 
    \label{mmsn_bcrit_bcool}}   
\end{figure}

@@

\subsection{VSI in the MMSN}
To obtain an overall picture of the relevance of the VSI to the MMSN,
we calculate characteristic growth times $t_\mathrm{grow} = 1/\nu$ of 
the fundamental mode in units of the local Keplerian orbital period
$P_\mathrm{orb}= 2\pi/\Omega_k$. The results are shown in 
Fig. \ref{mmsn_overall} for perturbation wavenumbers $5\lesssim
\khat\lesssim 50$. We consider this range because for  
$\khat=O(1)$ or $\khat=O(10^2)$ we found eigenfunctions have large
amplitudes near the vertical boundaries, so such modes may be affected
by boundary conditions.

>>>>

%In our disk models we have adopted a simple thermal relaxation 
%function in the energy equation. 
For a realistic disk one can consider the energy equation with
radiative cooling, 
%To connect our model with a realistic
%disk, we consider a more physically-motivated form of the energy
%equation with radiative cooling 
\begin{align}\label{real_energy}
%\rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
  c_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}where the temperature $T$ is defined through $P=\mathcal{R}\rho
T/\mu$, $\mathcal{R}$ is the ideal gas constant and $\mu$ is
the mean molecular weight, $c_v$ is the heat capacity at constant
volume, 
% In Eq. \ref{real_energy} $S$ is the
% specific entropy,
% \begin{align}
%   S \equiv c_v\ln{\left(\frac{p}{\rho^{\gamma}}\right)} 
% \end{align}
% where $c_v \equiv \left(\gamma-1\right)^{-1}\mathcal{R}/\mu$ is the
% specific heat capcity at constant volume
$\bm{F}$ is the radiative flux, and $Q_+$ is a heating
term such that an equilibrium solution exists.

One should now model in detail each of the terms on the right
hand side (RHS) of Eq. \ref{real_energy}, linearize them, then solve
the resulting linear eigenvalue problem. Such an approach is beyond
the scope of this paper. Here, we simply model the RHS of
Eq. \ref{real_energy} such that it can be represented by a constant
thermal relaxation time $t_c$ as used in our linear calculations.  
%Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
%the 
To do so, we first consider separately perturbations
with radial lengthscales $l\equiv 2\pi/k_x\gg l_\mathrm{rad}$ and 
$l\ll l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity.

%need to comment on kappa_d(rho,T)

\subsection{Radiative diffusion}
Let us first consider perturbations with lengthscales $l\gg
l_\mathrm{rad}$. In this case, we assume  
temperature fluctuations are smoothed out by radiative diffusion. We
mode this in linear theory by writing the linearized RHS of
Eq. \ref{real_energy} as 
% we choose the heating term such that
% \begin{align}
%   Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
%     \left(T-T_{t=0}\right)\right], 
% \end{align} 
% where $k_\mathrm{rad}$ is the conduction coefficient to be defined
% implicitly. Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
%     c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
% \end{align}
% When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right),  
\end{align}
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly.

However, Eq. \ref{diff_cool_proper} fundamentally changes the linear problem by
introducing higher order derivatives. We defer a proper analysis of
this problem to a follow-up study. Here, we are interested in
order-of-magnitude estimates for perturbations with radial
lengthscales much smaller than vertical lengthscales. We thus proceed
by assuming vertical derivatives can be neglected in
Eq. \ref{diff_cool_proper} and write  
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
% we are interested in radially-thin perturbations  
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}, 
\end{align}
where $\sigma_s$ is the Stefan-Boltzmann constant. 
From Eq. \ref{diff_cool_approx} we identify the scale-dependent thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}
where $\hat{\eta} = \eta/H^2\Omega_k$ is the 
dimensionless diffusion coefficient.

\subsection{Newtonian cooling}\label{newton_cool}
For perturbations with small lengthscales, $l\ll l_\mathrm{rad}$, 
the disk is optically thin. In this case, we assume 
Newtonian cooling can be applied, which is the thermal
relaxation model as adopted in our linear models. That is, temperature
flucations $\delta T$ decay on a timescale $t_\mathrm{cool}$,
independent of $k_x$. For simplicity, we take  
% the radiative cooling is given by {\bf (reference?)}
% \begin{align}
% \nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
% \end{align}   
% where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
% heating term 
% \begin{align}
%   Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
% \end{align}
% Then Eq. \ref{real_energy} becomes
% \begin{align}
%   \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
%   \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right), 
% \end{align}
% where $c_v$ is the heat capacity at constant volume. When linearized, the last term on the RHS becomes
% \begin{align}
%   -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
%     \frac{P}{\rho}\delta\rho\right). 
% \end{align}
% By comparing this expression with the linearized version of the energy
% equation used in our calculations (Eq. \ref{full_energy1}), we identify 
% \begin{align}\label{tc_newton_cool} 
%   t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
% \end{align}
% as the thermal relaxation timescale for optically thin cooling. 
\begin{align}
t_\mathrm{cool} = \frac{l_\mathrm{rad}^2}{3\eta}. 
\end{align}

\subsection{A model for combined thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
 % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
%where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
%$\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
%and Eq. \ref{eta_def}. 
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$.

In order to utilize our linear results, which assumes $t_c$ is a
constant, we need to make further 
approximations. Specifically, we consider vertically-isothermal disks
and replace explicit $\rho$ dependencies in the above expressions by $\Sigma/
H$, where $\Sigma$ is the surface density at the fiducial
radius of interest. For example, the dimensionless diffusion
coefficient becomes 
\begin{align}\label{diff_coeff_dimensionless}
  \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
    c_v\Omega_k}. 
\end{align}
Furthermore, we will choose a temperature-dependent opacity model for
simplicity, so that $\kappa_d=\kappa_d(T)$. The dimensionless thermal
relaxation time $\beta$  
\begin{align}\label{real_beta}
  \beta \equiv t_c\Omega_k =
  \frac{1}{\hat{\eta}}\left(\frac{l_\mathrm{rad}^2}{3H^2}
    + \frac{1}{\khat^2}\right)  \simeq
  \frac{1}{\hat{\eta}}\left(\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{1}{\khat^2}\right)
\end{align}
is then a constant for vertically-isothermal disks, 
and we can directly apply previous results.

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $r_\mathrm{AU}\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)r_\mathrm{AU}^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}r_\mathrm{AU}^{33/7}+\frac{1}{\khat^2}\right).    
\end{align}
We refer to the case with the scalings
$\hat{\Sigma}=\hat{\kappa}_d=\hat{T}=1$ as the Minimum Mass Solar
Nebulae (MMSN).  We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is necessary for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
%becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} with with $\mu =2.33$ and $\gamma=1.4$.  We set
the scaling $\hat{T}=\hat{\Sigma}=1$ and consider three opacity scales:
$\hat{\kappa}_d=0.1, \,1 $ and $10$.

The figure suggests that the fundamental VSI 
operates in the outer parts of the MMSN at radii on 
the order of tens of AU ($\hat{\kappa}_d=1$, middle panel). 
However, the thermal relaxation time $\beta$
is not much shorter than $\beta_\mathrm{crit}$, so the VSI may be 
weak. Lowering the opacity shifts the curves
inward, but raises also them. In fact, lowering the opacity by an
order of magnitude renders the fiducial  disk stable to the fundametal
mode for the range of wavenumbers considered ($\hat{\kappa}_d=0.1$,
upper panel). Raising the opacity, on
the other hand, decreases the thermal timescales, but also shifts the
curves outward, so the fundamental mode is restricted to the outer
disk ($\hat{\kappa}_d=10$, lower panel).

%We remark that it is possible to satisfy $\beta<\beta_\mathrm{crit}$
%at small radii by choosing a sufficiently large $\khat$. However, our
%linear calculations suggest growth rates eventually decay as 
%$\khat\to\infty$. Thus, very small-scale VSI in the inner disk is
%probably unimportant.

\begin{figure}
  \includegraphics[scale=.43,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/bcrit_mmsn0d1} 
  \includegraphics[scale=.43,clip=true,trim=0cm 1.8cm 0cm 1cm]{figures/bcrit_mmsn}  
  \includegraphics[scale=.43,clip=true,trim=0cm 0cm 0cm 1cm]{figures/bcrit_mmsn10} 
  \caption{Dimensionless thermal relaxation timescales $\beta$ in a
    fiducial protoplanetary disk model. Eq. \ref{beta_mmsn} is plotted
    for three values of the 
    perturbation radial wavenumber: $\khat=1$ (dotted), $\khat=10$
    (dashed) and $\khat=100$ (dashed-dot), for three values of the
    opacity relative to the MMSN: $\hat{\kappa}_d=0.1$ (top),
    $\hat{\kappa}_d=1$ (middle) and $\hat{\kappa}_d=10$ (bottom).  
    The solid line is the
    critical thermal relaxation timescale $\beta_\mathrm{crit}$. 
    % and 
    % $\beta<\beta_\mathrm{crit}$ is necessary for the fundamental isothermal VSI to operate  
    % for $\khat\gg1$. 
    \label{mmsn_bcrit_bcool}}   
\end{figure}

We remark that it is possible, in principle, to satisfy $\beta <
\beta_\mathrm{crit}$ by choosing a sufficiently large $\khat$,
equivalently a sufficiently small perturbation lengthscales. We
illustrate this by plotting at each radius the value of $\khat$ such
that $\beta = \beta_\mathrm{crit}$. We see that in order for the
fundamental VSI to operate in the inner disk (few AU), one needs to
consider large values of $\khat$.

However, our linear
calculations indicate growth rates eventually decay as one increases
$\khat$. This, together with our finding of high $\khat$ modes may be
affected by vertical boundary conditions, suggest the VSI does not
operate in the inner few AU in protoplanetary disks where thermal
relaxation is entirely due to radiation \citep[cf.][]{stoll14}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/bcrit_mink} 
  \caption{The minimum value of the perturbation wavenumber $\khat$ in
    the MMSN such that the associated dimensionless thermal
    relaxation time  $\beta$ is  less than the critical timescale
    $\beta_\mathrm{crit}$. 
    \label{mmsn_bcrit_bcool_mink}}   
\end{figure}

@@

\subsection{VSI in the outer parts of the MMSN}
To obtain an overall picture of the relevance of the VSI to the MMSN,
we calculate characteristic growth times $t_\mathrm{grow} = 1/\nu$ of 
the fundamental mode in units of the local Keplerian orbital period
$P_\mathrm{orb}= 2\pi/\Omega_k$. The results are shown in 
Fig. \ref{mmsn_overall} for perturbation wavenumbers $5\lesssim
\khat\lesssim 50$. We consider this range because for  
$\khat=O(1)$ or $\khat=O(10^2)$ we found eigenfunctions have large
amplitudes near the vertical boundaries, so such modes may be affected
by boundary conditions.
